---
title: "Reproducible Research - PA1"
author: "V. Kirilenko"
date: "Tuesday, March 11, 2015"
output: html_document
keep_md: true
---
# "Reproducible Research - PA1"   
**author: V.Kirilenko**
```{r "setoptions", echo = F}
library(knitr)
opts_chunk$set(echo = T, cache = F)
time = format(Sys.time(), "%X %d/%m/%Y")
```
  
  
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

### _Loading and preprocessing the data_
```{r "Download a code"}
url1 = "http://d396qusza40orc.cloudfront.net/repdata/data/activity.zip"
temp = tempfile()
download.file(url1, temp)
activity = read.csv(unz(temp, "activity.csv"))
unlink(temp)
```
  
  
### First question: what is mean total number of steps taken per day?

**_1. Plot total number of steps taken each day and calculate mean & median_**  
```{r "Plotting steps data", fig.height=6, fig.width=9}
steps = aggregate(activity$steps, by=list(activity$date), sum, na.rm=TRUE)
names(steps) = c("Date","Sum_of_steps")
hist(steps$Sum_of_steps,breaks = 30, freq = T, col = "darkred", xlab = "Sum of steps",
     main = paste("Total number of steps taken each day"))
```

**_2. Calculate and report the mean and median total number of steps taken per day_**  
``` {r }
matrix(c(mean(steps$Sum_of_steps),median(steps$Sum_of_steps)),nrow = 1, ncol = 2,
       dimnames = list(("Steps"),c("Mean","Median")))
```
  
  
### Second question: what is the average daily activity pattern?  
**_1. Plot means of steps by interval_**  
So, there are the gaps between 55 min and 00 min.  
Program get it as 45-length interval and plot interpolated line from the first value to the second value.  
To avoid this, I created an extra column, which has the intervals converted to time intervals.  
``` {r "Calculating mean of steps by interval and plot means by interval", fig.height=6, fig.width=9}
l = length(activity[,3])
for(i in 1:l){
  while (nchar(activity[i,3])<4)
    activity[i,3] = paste(as.character(0),as.character(activity[i,3]),sep ="")}
interval = aggregate(activity$steps, by=list(activity$interval), mean, na.rm=TRUE)
interval$Time = strptime((interval[,1]), format = "%H%M")
names(interval) = c("Interval","Mean","Time")
max_interval = (interval[which(grepl(max(interval$Mean), interval$Mean)),3])
plot(interval$Time, interval$Mean, type="l", xlab="Interval", xaxt="n", ylab="Mean of steps",
     main = paste("Average daily activity pattern"))
abline(v=axis.POSIXct(1, at = max_interval,col = "red",lwd = 2),col = "gray", lty = "dotted",lwd = par("lwd"))
axis.POSIXct(side = 1, at = c(interval[37,3],interval[85,3], interval[133,3],interval[169,3], 
                              interval[217,3],interval[265,3]))  
```  
  
The 5-minute interval that contains, in average, the maximum number of steps is **`r as.character((max_interval), format = "%H:%M")`.**  
  
### Third question: imputing missing values  
**_1. Calculating sum of NA in dataset_**  
``` {r "Calculating sum of NA values"}
sum(is.na(activity))
```

**_2. Imputing NA in dataset_**   
I choose the median of steps as data for replacing data NA values.  
I think that median is more stable with deviations than the mean.  
So, I got medians of steps each day and exchanged them with NA values where they were present.  
``` {r "Make NA`s replacement"}
act_temp = aggregate(steps ~ interval, data=activity, median)
names(act_temp)[2] = "medianSteps"
new_act = merge(activity, act_temp)
new_act$steps[is.na(new_act$steps)] = new_act$medianSteps[is.na(new_act$steps)]
median_steps = aggregate(new_act$steps, by=list(new_act$date), sum)
names(median_steps) = c("Date","Sum_of_steps")
```

**_3. Plot new dataset and calculating new mean & median_**  
``` {r "Plot new hist with imputing data"}
hist(median_steps$Sum_of_steps,breaks = 30, freq = T, col = "darkred", xlab = "Sum of steps",
     main = paste("Total number of steps taken each day with replacement of NA"))
matrix(c(c(mean(steps$Sum_of_steps), mean(median_steps$Sum_of_steps)),c(median(steps$Sum_of_steps), 
 median(median_steps$Sum_of_steps))),nrow = 2, ncol = 2, dimnames = list(c("Steps with NA","Steps w/o NA"),
                                                                         c("Mean","Median")))
```

**_4. Also we can check sums of steps in whole datasets:_**  
```{r "Calculate and print ratio between steps in new and old datasets"}
old_sum = sum(activity$steps,na.rm = T)
new_sum = sum(new_act$steps)
ratio = paste(format(round(((new_sum/old_sum)-1)*100, 2), nsmall = 2),"%")
matrix(c(old_sum,new_sum),nrow = 1, ncol = 2,
       dimnames = list(("Sum of steps"),c("Old","New")))
```
As we can see, impact of imputing missing data on the total daily number of steps is very small.  
Difference between sums of steps (and means of steps also) is only **`r ratio`**.  
  
  
### Fourth question: are there differences in activity patterns between weekdays and weekends?  
**_1. Calculate differences in activity patterns and merge the data_**  
``` {r "Split and combine data by day of week"}
activity$Date_Time = strptime(paste(activity$date, activity$interval), format = "%Y-%m-%d %H%M")
activity$weekdays = weekdays(activity[,4])
weekend = subset(activity, activity$weekdays %in% c("Saturday","Sunday"))
weekday = subset(activity, !(activity$weekdays %in% c("Saturday","Sunday")))
intWE = aggregate(weekend$steps, by=list(weekend$interval), mean, na.rm=TRUE)
intWE$day = "weekend"
intWD = aggregate(weekday$steps, by=list(weekday$interval), mean, na.rm=TRUE)
intWD$day = "weekday"
all_data = rbind(intWE,intWD)
names(all_data) = c("Interval","Mean","Day")
all_data$Mean[all_data$Day == "weekend"] = -all_data$Mean[all_data$Day == "weekend"]
```

**_2. Plot difference by types of days_**  
``` {r "Plot combined dataset", fig.height=6, fig.width=9}
library("ggplot2")
ggplot(all_data, aes(group = Day,Interval,Mean, col = Day)) + geom_line() + 
  ggtitle(expression(atop(bold("Differences in activity patterns between weekdays and weekends")))) +
  theme(plot.title=element_text(color="blue"),legend.position="bottom")+ scale_colour_hue(name = "Day
  of week:") + theme(axis.text=element_text(size=12), axis.title=element_text(size=14,face="bold")) +
  scale_x_discrete(breaks=c("0100", "0300", "0500", "0700", "0900", "1100", "1300",
                            "1500", "1700", "1900", "2100", "2300"))

```
  
**Last edition: `r time`**